Fractional Types

### 2019-11-22 fracGC/p.tex
abramsky's pic
leave ednote in comments only
Make Generalized Toffoli fit in 1 column. Saves a lot of space...
done with 6.2 for now

The denotation of the fractional type is now:

@@

Combinators:

@@

\begin{proposition}
sing is a monad; comonad; etc
\end{proposition}
\begin{proof}
\PIPFCombderive{}
\end{proof}

  properties of singleton, recip
derive join etc
monad,comonad structure

Evaluation:
\PIPFeval{}

Combine with eval, the whole thing is reversible:
\PIPFrev{}

show a small example

@@

@@

\PPFTname{}

@@

\PPFThof{}

>>>>

Following the development for the toy language above, we create a
lifted version of fractional $\Pi$ with pointed types. In addition to
the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
multiplicative structure over pointed types, fractional types, and a
special kind of pointed type that includes just one value, a singleton
type. The singleton types will allow the type system to track the flow
of one particular value, which is exactly what is needed to prove the
safety of deallocation:

@@

As can be seen, a lifted type is interpreted as a type along with a
corresponding value. A singleton type consists a special value of a
given type: any claim that a value belongs to the singleton type must
come with a proof that this value is equivalent to the special
value. A fractional type consumes a particular singleton type
returning the unit type. Note that the runtime gc process simply
disregards the value it is given: the type system however prevents
this value from being applied to anything but the particular singleton
value in question.

The combinators in the lifted language consist of all the combinators
in the base language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

The mediation between pointed and singleton types is based a dual
monad/comonad pair from which many structural properties can be
derived.

\begin{proposition}
  $\llparenthesis \cdot \rrparenthesis$ is both an idempotent strong
  monad and an idempotent costrong comonad over pointed types.
\end{proposition}
\begin{proof}
  The main insight needed is to define the functor \Afun{∙Singᵤ}, the
  tensor/cotensor, and the join/cojoin (duplicate):
\PIPFCombderive{}
\end{proof}

Like for the toy language, evaluation is reflected in the type
system, and in this case we have the additional property that
evaluation is reversible:

\PIPFeval{}

\PIPFrev{}

Thus, to summarize if a combinator expects a singleton type, then it
would only typecheck in the lifted language, if it is given the unique
value it expects. A particularly intringuing instance of that
situation is the following program:

@@

The program takes a value of type
\Acon{∙𝟙/}~(\Acon{∙𝟙/}~\Avar{A}). This would be a gc process
specialized to collect another gc process! At runtime type there would
be no information other than the functions that ignore their argument
but the type system provides enough guarantees to ensure that this
process is well-defined, is reversible, and is safe. The following
trace shows the actual execution for confirmation:

@@

thinking of gc as space transformers

\PPFTname{}
\PPFThof{}

trace about gc'ing cycles

@@

\includegraphics[scale=0.5]{trace.png}
